Real-Time Activity Classification by Matched Filtering using Body-Worn Accelerometers

### 2016-09-28 notes/paper/paper.tex
Incremental updates to the paper

\documentclass[journal]{IEEEtran}
\usepackage{blindtext}
\usepackage{graphicx}
\usepackage{subfig}
\usepackage[english]{babel}
\usepackage{pinyin}
\usepackage{natbib}
\usepackage{notoccite}
\usepackage{amssymb}
\usepackage{textcomp}
\usepackage{floatrow}
\usepackage{amsmath}
%
\begin{document}
%
% paper title
\title{Real-Time Activity Classification by Matched Filtering using Body-Worn Accelerometers}
%
\author{Craig~Euler, C.T.~Lin, Bryan~Juarez, Melissa~Flores}
%
\maketitle
%
\begin{abstract}
Having information of a user\textquotesingle s activity can provide great use in modern-day devices such as ones that track and monitor the user\textquotesingle s activity and fitness.
In this paper, we demonstrate activity classification performance of the matched-filtering method with data obtained from a three-axis accelerometer worn by the user.
We also show the real-time processing capability of our algorithms on the MSP432P401R microcontroller.
Dimensionality reduction with principal component analysis (PCA) \cite{bishop_2006} is a data compression technique we use to improve our processing throughput which, inherently, has the added benefit of making our data invariant to sensor orientation.
Data decimation in time is an additional throughput enhancement that we apply early to our data.
We make use of an instance-based learning algorithm to train the device to learn the individual\textquotesingle s motion patterns and store that information as activity templates for use in our matched-filter.
\end{abstract}
%
\section{Introduction}
Various methods for activity classification using body-worn sensors exist.
Custom Decision Tree, Automatically Generated Decision Tree, and Artificial Neural Networks \cite{parkka_ermes_korpipaa_mantyjarvi_peltola_korhonen_2006} have been explored as well as other frequency based classification methods \cite{sharma_purwar_lee_lee_chung_2008}.
In this paper, we make use of the matched-filter method to identify an individual\textquotesingle s activity for the purpose of real-time processing.
The matched-filter method with use of thresholding and comparing the correlations of various filters is a simple method to use and has been studied in the past \cite{giannakis_tsatsanis_1990}.
The methods used for our algorithms are designed for computational efficiency to accommodate the resource limitations of a microcontroller (we use the MSP432 microcontroller).

@@

The matched-filter is based on the cross correlation of the data $\textbf{y} = \{y\}_{k=0}^{N-1}$ with the reference signal $\textbf{x} = \{x\}_{k=0}^{M-1}$.
The cross correlation of $\textbf{x}$ with $\textbf{y}$ is defined as:
%
\begin{equation} \label{eq:cross_correlation}
(\textbf{x} \star \textbf{y}) := \left \{\sum_{i=0}^{M-1}x_{i} y_{i+k} \right \}_{k=0}^{N-M-1}
\end{equation}
%
which is simply a moving dot product.
We can then normalize \eqref{eq:cross_correlation} by the following:
%
\begin{equation} \label{eq:norm_cross_correlation}
\widehat{(\textbf{x} \star \textbf{y})}_k := \frac{(\textbf{x} \star \textbf{y})_k}{||\textbf{x}|| \ || \widetilde{\textbf{y}}_k || }, \quad k = 0,1,...,N-M-1
\end{equation}
%
where $|| \cdot ||$ is the usual $\ell_2$ norm and $\widetilde{\textbf{y}}_k = \{y_p\}_{p=k}^{M+k-1}$.
\eqref{eq:norm_cross_correlation} is now bounded by $[-1,1]$ and has any bias toward high energy removed.
We improve our computational efficiency by applying the convolution theorem to \eqref{eq:cross_correlation} and replacing it with:
%
\begin{equation} \label{eq:conv_theorem}
(\textbf{x} \star \textbf{y}) := FFT^{-1}(FFT(\textbf{x}') FFT(\textbf{y})^*)
\end{equation}
%
where $\textbf{x}'$ is $\textbf{x}$ that has been zero padded to the length of $\textbf{y}$ with the added restriction that the length of $\textbf{x}$ is no more than the length of $\textbf{y}$ and $FFT$ is the Fast Fourier Transform algorithm.
We are most interested where our reference $\textbf{x}$ best matches our data $\textbf{y}$.
We use the resulting max correlation to define our metric for  our matched-filter output as:
%
\begin{equation} \label{eq:matched_filter}
MF_{\textbf{x}}(\textbf{y}) := \underset{k \in [0, M-N)}{max} \left \{\widehat{(\textbf{x} \star \textbf{y})}_k \right \}
\end{equation}
%
If \eqref{eq:matched_filter} passes a set threshold, then the corresponding signal $\textbf{y}$ is classified as a match to the reference $\textbf{x}$.
%
\subsection{Dimensionality Reduction}
We manage to reduce our computational resources and remove our dependence on sensor orientation at the same time through a process of dimensionality reduction with principal component analysis (PCA) \cite{bishop_2006}.
We assume that over a short period of time, a user\textquotesingle s limb movement is constrained mostly to a two-dimensional plane.
By identifying this plane of motion, we then project the acceleration data onto this plane and re-define our new axis accordingly.

@@

Classification methods that are achieved from multi-dimensional feature spaces have been explored \cite{survey_multi_view_machine_learning}.
In this paper, we compute the matched-filter output for two dimensions by generalizing \eqref{eq:matched_filter} by augmenting our reference and signal with both dimensions appropriately for normalization and adding together the cross correlations as:
%
\begin{equation} \label{eq:cross_correlation_2}
\widehat{(\textbf{x}^{1,2} \star \textbf{y}^{1,2})}_k := \frac{(\textbf{x}^1 \star \textbf{y}^1)_k + (\textbf{x}^2 \star \textbf{y}^2)_k}{||[ \textbf{x}^1, \textbf{x}^2 ]|| \ || [ \widetilde{\textbf{y}}_k^1, \widetilde{\textbf{y}}_k^2 ] || },
\end{equation}
%
for $ k = 0,1,...,N-M-1 $. \\
%
\begin{equation} \label{eq:matched_filter_2}
MF_{\textbf{x}}^2(\textbf{y}) := \underset{k \in [0, M-N)}{max} \left \{\widehat{(\textbf{x}^{1,2} \star \textbf{y}^{1,2})}_k \right \}
\end{equation}
%
This works because both dimensions are always in phase with each other for any given activity and therefore the match will happen in sync. Here, the superscripts $1$ and $2$ represent the primary and secondary principal components of our data respectively.
%
\subsection{Decimation}
Decimation is the first step applied to our data before any additional processing takes place.
Decimation is a process of reducing the sampling rate of the signal by filtering to mitigate any possible distortion effects due to aliasing followed by downsampling.
Because real-time performance is important, we decimate our signal for the purpose of improving our throughput.

@@

\begin{figure}[!ht]
   \centering
   \subfloat[\label{fig:multiple_comparison}]{\includegraphics[width=0.45\textwidth]{more_time_plots.png}}
   \quad
   \subfloat[\label{fig:timing_refs}]{\includegraphics[width=0.45\textwidth]{one_and_two_references.png}}
   \caption{Timing results with varying sampling frequencies, time-window lengths, and number of references used for the matched-filter on the MSP432 microcontroller processor. The time window used for (b) is 4.0 seconds with timings measured at sampling frequencies of $16, 32, 62, 64, 75, 100,$ and $128$ Hz.}
   \label{fig:timing}
\end{figure}
%
Because the accelerometer was not connected to the microcontroller at the time of writing this paper, randomized data was used for the purpose of analyzing the throughput performance on the device while varying the parameters which accurately models the processing load we would expect when we do have the accelerometer connected to the microcontroller.
Figure \ref{fig:multiple_comparison} shows the execution time with varying time-window lengths and sampling frequencies while maintaining two references used in the matched-filter process and operating at decimation $2$.
Because the FFT dominates the processing load, and a custom FFT was written specifically for this unit and optimized for even-powered factors, one expects best throughput at radix 2 valued time-window lengths and sampling frequencies followed by other even-valued, non-radix 2, time-lengths and sampling frequencies. Both $62$Hz and $50$Hz have only one factor of $2$ hence the expectation that a sampling frequency of $62$Hz processing time is longer than that of the $50$Hz.
Figure \ref{fig:timing_refs} shows the execution time over a $4.0$ second time window at decimation $8$ while varying the sampling frequency and number of references. Because the pre-processing execution time does not change with the number of references, it's observed that matched-filtering accounts for roughly $80$\% of the processing while pre-processing accounts for the other $20$\% when processing for a single reference.
%
\begin{figure}[!ht]
   \centering
   \includegraphics[width=0.95\textwidth]{flops_graph.png}
   \caption{An estimate of the processing distribution based on flop counts. This estimate was obtained from the source code that was used for the performance assessment. The cost of the FFT is based on the $2.5 N log_2 N$ mflops estimate for FFTW \cite{FFTW} which was used for this performance assessment for real-FFTs. A sampling rate of $128$ Hz with the matched-filter applied over a $4$ second time window and $1.5$ seconds of reference data was used (usual standard for testing)}
   \label{fig:flops_count}
\end{figure}

An assessment of the processing costs associated with each algorithm in the process was made as seen in figure \ref{fig:flops_count}.
In addition, timing tests were done for comparison on an Intel i7-4500U 1.80 GHz processor.
Processing time came out to $12.0$ micro-seconds for dimensionality reduction processing and $114.0$ micro-seconds for matched-filter processing on the remaining two dimensions.
This proportion shows relative consistency with figure \ref{fig:flops_count} as we would expect.
FFT calls account for the bulk of the processing distribution and are part of the main processing kernel for the matched-filtering process.
%
\section{Conclusion}
We have demonstrated excellent activity classification capability of the matched-filter method using data obtained from a Kinetisense tri-axis accelerometer sensor.
We have shown that real-time processing can be achieved with timing results obtained from the MSP432 microcontroller.
Timing results have demonstrated that radix 2 time-window lengths and frequency values result in faster processing times as expected given that the FFT dominates the processing time.
We increased our throughput with the use of data-compression techniques, namely decimation and dimensionality reduction, which reduced our computational requirements.
The references used in the matched-filter are obtained from the training module where a training set is provided by the user through some specified activity and the best representative motion signature of the individual is extracted through the use of an instance-based learning algorithm.

>>>>

\documentclass[conference]{IEEEtran}
\usepackage{blindtext}
\usepackage{graphicx}
\usepackage{subfig}
\usepackage[english]{babel}
\usepackage{pinyin}
\usepackage{natbib}
\usepackage{notoccite}
\usepackage{amssymb}
\usepackage{textcomp}
\usepackage{floatrow}
\usepackage{amsmath}
%
\begin{document}
%
% paper title
\title{Real-Time Activity Classification by Matched Filtering using Body-Worn Accelerometers}
%
\author{Craig~Euler craig.euler@gmail.com, C.T.~Lin ct.lin@csun.edu,\\Bryan~Juarez bryan.juarez.974@my.csun.edu, Melissa~Flores melissa.flores.455@my.csun.edu\\College of Engineering and Computer Science\\California State University, Northridge}
%
\maketitle
%
\begin{abstract}
Having information of a user\textquotesingle s activity can provide great use in modern-day devices such as ones that track and monitor the user\textquotesingle s activity and fitness.
In this paper, we demonstrate activity classification performance of the matched-filtering method with data obtained from a three-axis accelerometer worn by the user.
We also show the real-time processing capability of our algorithms on the MSP432P401R microcontroller.
Dimensionality reduction with principal component analysis (PCA) \cite{bishop_2006} is a data compression technique we use to improve our processing throughput which, inherently, has the added benefit of making our data invariant to sensor orientation.
Data decimation in time is an additional throughput enhancement that we apply early to our data.
We make use of an instance-based learning algorithm to train the device to learn the individual\textquotesingle s motion patterns and store that information as activity templates for use in our matched-filter.
\end{abstract}
%
\section{Introduction}
Various methods for activity classification using body-worn sensors exist.
Custom Decision Tree, Automatically Generated Decision Tree, and Artificial Neural Networks \cite{parkka_ermes_korpipaa_mantyjarvi_peltola_korhonen_2006} have been explored as well as other frequency based classification methods \cite{sharma_purwar_lee_lee_chung_2008}.
In this paper, we make use of the matched-filter method to identify an individual\textquotesingle s activity for the purpose of real-time processing.
The matched-filter method with use of thresholding and comparing the correlations of various filters is a simple method to use and has been studied in the past \cite{giannakis_tsatsanis_1990}.
The methods used for our algorithms are designed for computational efficiency to accommodate the resource limitations of a microcontroller (we use the MSP432 microcontroller).

@@

The matched-filter is based on the cross correlation of the data $\textbf{y} = \{y\}_{k=0}^{N-1}$ with the reference signal $\textbf{x} = \{x\}_{k=0}^{M-1}$.
The cross correlation of $\textbf{x}$ with $\textbf{y}$ is defined as:
%
\begin{equation} \label{eq:cross_correlation}
(\textbf{x} \star \textbf{y}) := \left \{\sum_{i=0}^{M-1}x_{i} y_{i+k} \right \}_{k=0}^{N-M-1}
\end{equation}
%
which is simply a moving dot product.
We can then normalize \eqref{eq:cross_correlation} by the following:
%
\begin{equation} \label{eq:norm_cross_correlation}
\widehat{(\textbf{x} \star \textbf{y})}_k := \frac{(\textbf{x} \star \textbf{y})_k}{||\textbf{x}|| \ || \widetilde{\textbf{y}}_k || }, \quad k = 0,1,...,N-M-1
\end{equation}
%
where $|| \cdot ||$ is the usual $\ell_2$ norm and $\widetilde{\textbf{y}}_k = \{y_p\}_{p=k}^{M+k-1}$.
\eqref{eq:norm_cross_correlation} is now bounded by $[-1,1]$ and has any bias toward high energy removed.
We improve our computational efficiency by applying the convolution theorem to \eqref{eq:cross_correlation} and replacing it with:
%
\begin{equation} \label{eq:conv_theorem}
(\textbf{x} \star \textbf{y}) := FFT^{-1}(FFT(\textbf{x}') FFT(\textbf{y})^*)
\end{equation}
%
where $\textbf{x}'$ is $\textbf{x}$ that has been zero padded to the length of $\textbf{y}$ with the added restriction that the length of $\textbf{x}$ is no more than the length of $\textbf{y}$ and $FFT$ is the Fast Fourier Transform algorithm.
We are most interested where our reference $\textbf{x}$ best matches our data $\textbf{y}$.
We use the resulting max correlation to define our metric for  our matched-filter output as:
%
\begin{equation} \label{eq:matched_filter}
MF_{\textbf{x}}(\textbf{y}) := \underset{k \in [0, M-N)}{max} \left \{\widehat{(\textbf{x} \star \textbf{y})}_k \right \}
\end{equation}
%
If \eqref{eq:matched_filter} passes a set threshold, then the corresponding signal $\textbf{y}$ is classified as a match to the reference $\textbf{x}$.
%
\subsection{Dimensionality Reduction}
We manage to reduce our computational resources and remove our dependence on sensor orientation at the same time through a process of dimensionality reduction with principal component analysis (PCA) \cite{bishop_2006}.
Similar to other methods that apply subspace projections \cite{low_rank_subspace}, we seek to analyze our data in a lower dimensional subspace by a projection onto the two-dimension plane that best represents our data.
We assume that over a short period of time, a user\textquotesingle s limb movement is constrained mostly to a two-dimensional plane.
By identifying this plane of motion, we then project the acceleration data onto this plane and re-define our new axis accordingly.

@@

Classification methods that make use of multi-dimensional feature spaces have been explored \cite{survey_multi_view_machine_learning}.
In this paper, we compute the matched-filter output for two dimensions by generalizing \eqref{eq:matched_filter} by augmenting our reference and signal with both dimensions appropriately for normalization and adding together the cross correlations as:
%
\begin{equation} \label{eq:cross_correlation_2}
\widehat{(\textbf{x}^{1,2} \star \textbf{y}^{1,2})}_k := \frac{(\textbf{x}^1 \star \textbf{y}^1)_k + (\textbf{x}^2 \star \textbf{y}^2)_k}{||[ \textbf{x}^1, \textbf{x}^2 ]|| \ || [ \widetilde{\textbf{y}}_k^1, \widetilde{\textbf{y}}_k^2 ] || },
\end{equation}
%
for $ k = 0,1,...,N-M-1 $. \\
%
\begin{equation} \label{eq:matched_filter_2}
MF_{\textbf{x}}^2(\textbf{y}) := \underset{k \in [0, M-N)}{max} \left \{\widehat{(\textbf{x}^{1,2} \star \textbf{y}^{1,2})}_k \right \}
\end{equation}
%
This works because both dimensions are always in phase with each other for any given activity and therefore the match will happen in sync. Here, the superscripts $1$ and $2$ represent the primary and secondary principal components of our data respectively.
%
\subsection{Decimation}
Decimation is the first step applied to our data before any additional processing takes place.
Decimation is a process of reducing the sampling rate of the signal by filtering to mitigate any possible distortion effects due to aliasing followed by downsampling.
Because real-time performance is important, we decimate our signal for the purpose of improving our throughput.

@@

\begin{figure}[!ht]
   \centering
   \subfloat[\label{fig:multiple_comparison}]{\includegraphics[width=0.45\textwidth]{more_time_plots.png}}
   \quad
   \subfloat[\label{fig:timing_refs}]{\includegraphics[width=0.45\textwidth]{one_and_two_references.png}}
   \caption{Timing results with varying sampling frequencies, time-window lengths, and number of references used for the matched-filter on the MSP432 microcontroller processor. The time window used for (b) is 4.0 seconds with timings measured at sampling frequencies of $16, 32, 62, 64, 75, 100,$ and $128$ Hz.}
   \label{fig:timing}
\end{figure}
%
Because the accelerometer was not connected to the microcontroller at the time of writing this paper, randomized data was used for the purpose of analyzing the throughput performance on the device while varying the parameters which accurately models the processing load we would expect when we do have the accelerometer connected to the microcontroller.
Figure \ref{fig:multiple_comparison} shows the execution time with varying time-window lengths and sampling frequencies while maintaining two references used in the matched-filter process and operating at decimation $2$.
Because the FFT dominates the processing load, and a custom FFT was written specifically for this unit and optimized for even-powered factors, one expects best throughput at radix 2 valued time-window lengths and sampling frequencies followed by other even-valued, non-radix 2, time-lengths and sampling frequencies. Both $62$Hz and $50$Hz have only one factor of $2$ hence the expectation that a sampling frequency of $62$Hz processing time is longer than that of the $50$Hz.
Figure \ref{fig:timing_refs} shows the execution time over a $4.0$ second time window at decimation $8$ while varying the sampling frequency and number of references. Because the pre-processing execution time does not change with the number of references, it's observed that matched-filtering accounts for roughly $80$\% of the processing while pre-processing accounts for the other $20$\% when processing for a single reference.
%
\begin{figure}[!ht]
   \centering
   \includegraphics[width=0.99\textwidth]{flops_graph.png}
   \caption{An estimate of the processing distribution based on flop counts. This estimate was obtained from the source code that was used for the performance assessment. The cost of the FFT is based on the $2.5 N log_2 N$ mflops estimate for real-FFTs from FFTW \cite{FFTW} which was used for this performance assessment. A sampling rate of $128$ Hz with the matched-filter applied over a $4$ second time window and $1.5$ seconds of reference data was used (usual standard for our testing).}
   \label{fig:flops_count}
\end{figure}

An assessment of the processing costs associated with each algorithm in the process was made as seen in figure \ref{fig:flops_count}.
In addition, timing tests were done for comparison on an Intel i7-4500U 1.80 GHz processor.
Processing time came out to $12.0$ micro-seconds for dimensionality reduction processing and $114.0$ micro-seconds for matched-filter processing on the remaining two dimensions.
This proportion shows relative consistency with figure \ref{fig:flops_count} as we would expect.
FFT calls account for the bulk of the processing distribution and are part of the main processing kernel for the matched-filtering process.
%
\section{Conclusion}
We have demonstrated excellent activity classification capability of the matched-filter method using data obtained from a Kinetisense tri-axis accelerometer sensor.
We have shown that real-time processing on a low-powered micro-controller can be achieved with timing results obtained from the MSP432 microcontroller.
Timing results have demonstrated that radix 2 time-window lengths and frequency values result in faster processing times as expected given that the FFT dominates the processing time.
We increased our throughput with the use of data-compression techniques, namely decimation and dimensionality reduction, which reduced our computational requirements.
The references used in the matched-filter are obtained from the training module where a training set is provided by the user through some specified activity and the best representative motion signature of the individual is extracted through the use of an instance-based learning algorithm.
